---
title: "Housing Price Growth Regression Analysis"
author: "Yarida Kaewthong"
date: "2026-05-15"
output:
  pdf_document: default
  html_document: default
---

```{r package-setup, include=FALSE}
# ------------------------------------------------------------
# IMPORTANT: Run this package setup first before running all code
# This section checks whether the required package is installed.
# If the package is missing, R will install it automatically.
# The package is needed to create the graphs in this analysis.
# ------------------------------------------------------------

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

required_packages <- c("ggplot2")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(ggplot2)
```
# 1. Load and Prepare Data

# Load dataset
# IMPORTANT: data_new.csv must be in the same folder as this RMarkdown file

```{r}
data <- read.csv("data_new.csv", check.names = FALSE, row.names = NULL)

names(data)

data_clean <- data[, c(
  "Year",
  "QuarterDate",
  "Quarter",
  "City",
  "State",
  "HPGrowth_pct(%)",
  "InterestRate(%)",
  "Inflation(%)",
  "NetOverseasMigration",
  "HouseholdIncome_per_capita",
  "Covid19",
  "Election_dummy",
  "Quarter2",
  "Quarter3",
  "Quarter4",
  "Time Trend"
)]

data_clean$State <- as.factor(data_clean$State)
data_clean$State <- relevel(data_clean$State, ref = "SA")

head(data_clean)
```
# 2. Regression Models

## Model 1: Baseline Time-Control Model

This model includes time trend, quarterly seasonality, and state controls.

```{r}
model1 <- lm(
  `HPGrowth_pct(%)` ~ Quarter2 + Quarter3 + Quarter4 + `Time Trend` + State,
  data = data_clean
)

summary(model1)
```

## Model 2: Adding Interest Rate and Inflation

This model adds basic macroeconomic variables.

```{r}
model2 <- lm(
  `HPGrowth_pct(%)` ~ `InterestRate(%)` + `Inflation(%)` +
    Quarter2 + Quarter3 + Quarter4 + `Time Trend` + State,
  data = data_clean
)

summary(model2)
```

## Model 3: Adding Migration and Household Income

This model adds net overseas migration and household income.

```{r}
model3 <- lm(
  `HPGrowth_pct(%)` ~ `InterestRate(%)` + `Inflation(%)` +
    NetOverseasMigration + HouseholdIncome_per_capita +
    Quarter2 + Quarter3 + Quarter4 + `Time Trend` + State,
  data = data_clean
)

summary(model3)
```

## Model 4: Adding COVID-19

This model adds the COVID-19 dummy variable.

```{r}
model4 <- lm(
  `HPGrowth_pct(%)` ~ `InterestRate(%)` + `Inflation(%)` +
    NetOverseasMigration + HouseholdIncome_per_capita + Covid19 +
    Quarter2 + Quarter3 + Quarter4 + `Time Trend` + State,
  data = data_clean
)

summary(model4)
```

## Model 5: Adding Election Dummy

This model adds the election dummy variable.

```{r}
model5 <- lm(
  `HPGrowth_pct(%)` ~ `InterestRate(%)` + `Inflation(%)` +
    NetOverseasMigration + HouseholdIncome_per_capita + Covid19 +
    Election_dummy + Quarter2 + Quarter3 + Quarter4 +
    `Time Trend` + State,
  data = data_clean
)

summary(model5)
```
# 3. Model Comparison

```{r}
get_f_pvalue <- function(model) {
  f <- summary(model)$fstatistic
  pf(f[1], f[2], f[3], lower.tail = FALSE)
}

comparison <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  R_Squared = c(
    summary(model1)$r.squared,
    summary(model2)$r.squared,
    summary(model3)$r.squared,
    summary(model4)$r.squared,
    summary(model5)$r.squared
  ),
  Adjusted_R_Squared = c(
    summary(model1)$adj.r.squared,
    summary(model2)$adj.r.squared,
    summary(model3)$adj.r.squared,
    summary(model4)$adj.r.squared,
    summary(model5)$adj.r.squared
  ),
  Significance_F = c(
    get_f_pvalue(model1),
    get_f_pvalue(model2),
    get_f_pvalue(model3),
    get_f_pvalue(model4),
    get_f_pvalue(model5)
  )
)

comparison
```

Model 4 is selected as the preferred model because it has the highest adjusted R-squared value. Although Model 5 has a slightly higher R-squared, its adjusted R-squared decreases after adding the election dummy variable. The election dummy is also not statistically significant.

# 4. Prediction Accuracy

```{r}
get_accuracy <- function(model, data) {
  predicted <- predict(model)
  actual <- data$`HPGrowth_pct(%)`
  residual <- actual - predicted
  
  rmse <- sqrt(mean(residual^2))
  mae <- mean(abs(residual))
  
  return(c(RMSE = rmse, MAE = mae))
}

accuracy_table <- data.frame(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5"),
  rbind(
    get_accuracy(model1, data_clean),
    get_accuracy(model2, data_clean),
    get_accuracy(model3, data_clean),
    get_accuracy(model4, data_clean),
    get_accuracy(model5, data_clean)
  )
)

accuracy_table
```

RMSE and MAE were calculated to compare prediction accuracy. Lower values indicate better predictive performance. MAPE was not used because housing price growth includes negative and near-zero values.

# 5. Actual vs Predicted Graphs

```{r}
data_clean$QuarterLabel <- ifelse(data_clean$Quarter2 == 1, "Q2",
                           ifelse(data_clean$Quarter3 == 1, "Q3",
                           ifelse(data_clean$Quarter4 == 1, "Q4", "Q1")))

data_clean$QuarterDate <- as.Date(
  paste(
    data_clean$Year,
    ifelse(data_clean$QuarterLabel == "Q1", "01-01",
    ifelse(data_clean$QuarterLabel == "Q2", "04-01",
    ifelse(data_clean$QuarterLabel == "Q3", "07-01", "10-01"))),
    sep = "-"
  )
)

data_clean$Predicted_Model4 <- predict(model4)
```

```{r}
plot_state <- function(state_name) {
  
  state_data <- subset(data_clean, State == state_name)
  
  ggplot(state_data, aes(x = QuarterDate)) +
    annotate(
      "rect",
      xmin = min(state_data$QuarterDate[state_data$Covid19 == 1]),
      xmax = max(state_data$QuarterDate[state_data$Covid19 == 1]),
      ymin = -Inf,
      ymax = Inf,
      alpha = 0.15,
      fill = "grey"
    ) +
    geom_line(aes(y = `HPGrowth_pct(%)`, colour = "Actual HPGrowth"), linewidth = 1) +
    geom_point(aes(y = `HPGrowth_pct(%)`, colour = "Actual HPGrowth"), size = 1.5) +
    geom_line(aes(y = Predicted_Model4, colour = "Predicted HPGrowth"),
              linewidth = 1, linetype = "dashed") +
    geom_point(aes(y = Predicted_Model4, colour = "Predicted HPGrowth"), size = 1.5) +
    scale_colour_manual(
      values = c(
        "Actual HPGrowth" = "steelblue",
        "Predicted HPGrowth" = "darkorange"
      )
    ) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(
      title = paste("Model 4: Actual vs Predicted Quarterly Housing Price Growth in", state_name),
      subtitle = "Each point represents one quarter; grey area represents the COVID-19 period",
      x = "Year",
      y = "Housing Price Growth (%)",
      colour = ""
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    )
}
```

## NSW

```{r, fig.width=11, fig.height=6, out.width="100%"}
plot_state("NSW")
```


## VIC

```{r, fig.width=11, fig.height=6, out.width="100%"}
plot_state("VIC")
```

## QLD

```{r, fig.width=11, fig.height=6, out.width="100%"}
plot_state("QLD")
```

## SA

```{r, fig.width=11, fig.height=6, out.width="100%"}
plot_state("SA")
```

## WA

```{r, fig.width=11, fig.height=6, out.width="100%"}
plot_state("WA")
```



# 6. Final Conclusion

The regression results show that Model 4 is the preferred model. It includes interest rate, inflation, net overseas migration, household income, COVID-19, quarterly seasonal controls, time trend, and state controls. Model 4 has the highest adjusted R-squared value before adding the election dummy. Model 5 has a slightly higher R-squared, but its adjusted R-squared decreases and the election dummy is not statistically significant. Therefore, Model 4 provides the best balance between explanatory power, prediction accuracy, and model simplicity.
